---
title: "Milestone 2"
author: "Abigail Duque"
date: "2022-10-02"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Description of datset

## What is the data source?

The dataset is from the European Statistical System/2011 European census data which includes the EU country codes, sex at birth, age ranges, employment status (CAS), education status, population of locality of residence and the number of people in each strata. 

## How does the dataset relate to the group problem statement and question?

The dataset will help us visualize how Monkeypox case rates differ by region, stratified by various demographic factors to prepare for an update to the state health department.

## Import Statement

## Document the import process

From the PHW251 Project Data Repository on Github, we opened each appropriate csv, obtained the url from download/viewing the raw data and then imported the csv into r studio using the read_csv function from the readr package.

```{r import csv}

## ECS = Euro Census Stats

library(readr)

url = 'https://raw.githubusercontent.com/PHW290/phw251_projectdata/main/euro_census_stats.csv'

ECS <- read_csv(url,
          col_names = TRUE,
          col_types = NULL,
          na = c("", "NA")
)
        

```

## Tidying dataframe
```{r tidying dataframe}

library(tidyverse)

ECS1 <- rename_with(ECS, ~ tolower(gsub(" ","_",.x,fixed=TRUE)))

```

## Identify data types for 5+ data elements/columns/variables


From read_csv, there are 8 character data types : COUNTRY_CODE, SEX, AGE, CAS, EDU, FLAGS, FOOTNOTES, RES_POP and 2 numeric data types:TIME, pop. <br>
We will obtain the different regions assessed from the country_code variable. The demographic variables we might use include sex, age, CAS (employment status) and EDU (education status). <br>
We will also need the population of locality of residence (RES_POP) and number of people in each strata (pop). <br>

This data set will merge with eu monkeypox cases & eu population denominators. The data types of each variable are acceptable.


## Provide a basic description of the 5+ data element


## variable: country_code
Using the unique function on the country_code variable, there are 32 different two-letter country codes in this dataset.
```{r country_code}

unique(ECS1$country_code)

```
## variable: sex
Two categories under sex: male, female.
```{r sex}
unique(ECS1$sex)

```

## variable: age
There are six different age ranges in the dataset: <15, 15-29, 30-49, 65-84 & >85
```{r age}

unique(ECS1$age)


```
## variable: cas
There are five different statuses of employment.
```{r employment status}

unique(ECS1$cas)

```
## variable: edu
There are nine different statuses of education.
```{r education status}

unique(ECS1$edu)


```

## variable: res_pop
There are seven levels of population in each locality of residence.
```{r residence population}

unique(ECS1$res_pop)


```

## variable: pop
The number of people in the different location strata ranges from 0 to 1,702,270.
```{r # of people in each location strata}

range(ECS1$pop)

```