---
title: "m2_pop_denoms"
author: "Katherine Boudreau"
output: pdf_document
date: '2022-10-02'
---

**in progress**

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse) #readr, dplyr, and tibble are included in this
```

Import Statement
```{r import}
pop_denoms = read_csv("https://raw.githubusercontent.com/PHW290/phw251_projectdata/main/euro_pop_denominators.csv")
```
Description of dataset
What is the data source? (1-2 sentences on where the data is coming from, dates included, etc.)
Eurostat, population data as of Jan 1
https://ec.europa.eu/eurostat/databrowser/view/tps00001/default/table?lang=en
The data source of euro_pop_denominators.csv

How does the dataset relate to the group problem statement and question?
The population data in this csv will serve as the denominator in the incident rate calculations.

```{r playground}
be <- pop_denoms[which (pop_denoms$geo=="BE"),]
```

Elements:
- geo is a string (char?) denoting the region that this population data comes from.
Geopolitical entity; 54 options

- OBS_VALUE is the observed population on January 1 of the reported year. Numeric.