---
title: "m2_pop_denoms"
author: "Katherine Boudreau"
output: pdf_document
date: '2022-10-02'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse) #readr, dplyr, and tibble are included in this
```

## Description of dataset

### What is the data source?
Our group took the European Population Denominators csv from the \href{https://github.com/PHW290/phw251_projectdata}{PHW251 Project Data github repo}. This data originates from \href{https://ec.europa.eu/eurostat/databrowser/view/tps00001/default/table?lang=en}{Eurostat} and includes European countries' yearly population data from 2011 to 2022, based on the total population residing in that country as of January 1st of each year.

### How does the dataset relate to the group problem statement and question?
This population data will serve as the denominator in the case rate calculations we report across various European regions and time periods.

## Import Statement
To import this csv, we copied the URL for the raw git hub view of the appropriate \href{https://github.com/PHW290/phw251_projectdata/blob/main/euro_pop_denominators.csv}{csv file}. We then imported the csv using the read_csv() readr function.
```{r import, message=FALSE}
file_path = "https://raw.githubusercontent.com/PHW290/phw251_projectdata/main/euro_pop_denominators.csv"
pop_denoms = read_csv(file_path)
```

## Data Types and Descriptions
Two key data columns from this file are the GEO column and the OBS_VALUE column.
```{r playground}
# GEO Column
summary(pop_denoms$geo)
typeof(pop_denoms$geo)

# OBS VALUE column
summary(pop_denoms$OBS_VALUE)
typeof(pop_denoms$OBS_VALUE)
```
The values in the GEO column are characters denoting the geopolitical region that a given row's population data comes from. There are 54 possible values for this column, either a 2-letter country code or a code denoting the entire European Union on a given year. We will eventually merge these 2-letter country codess with the 2-letter country codes in the euro_mpx_cases csv, so the character data type will suffice.

The values in the OBS_VALUE column are the observed population on January 1 of the reported year. These values are doubles representing the total population count of a given region, so we can use thess values directly as the denominators in our case rate calculations.