
---
title: "Milestone 4"
author: "Abigail Duque, Misty Garcia, Katherine Boudreau"
date: '2022-11-19'
output:html_document:
df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(tidyverse)
library(kableExtra)
library(plotly)
```

# Milestone 1
Github link: https://github.com/r-public-health-group/monkey-pox


Team roles and responsibilities: https://github.com/r-public-health-group/monkey-pox/blob/main/team_roles_and_responsibilities.docx

\newpage

# Milestone 2
## Description of dataset

### What is the data source?
We are interested in how monkeypox case rates differ by European region and population demographics. We are first utilizing a monkeypox case dataset from the European Centre for Disease Prevention and Control that spans 2022-05-09 to 2022-08-23 and includes the daily number of confirmed monkey pox cases by country and how the data was collected. Then, we will incorporate a European population dataset from \href{https://ec.europa.eu/eurostat/databrowser/view/tps00001/default/table?lang=en}{Eurostat}, which includes European countries' yearly population data from 2011 to 2022 based on the total population residing in that country as of January 1st of each year. Finally, we will incorporate the European Statistical System's 2011 European census data which includes the EU country codes, sex at birth, age ranges, employment status (CAS), education status, population of locality of residence and the number of people in each strata.

### How does the dataset relate to the group problem statement and question?
As we prepare our report on monkeypox case rates to the state health department, the monkeypox case data will provide us with the number of cases that will be the numerator of our case rate calculation, the population data gives us the size of the population at risk that will be the denominator of our case rate calculation, and the census data will allow us to stratify the case rates by various demographic factors.

## Import Statement
All three of our CSVs are hosted on the course github page: https://github.com/PHW290/phw251_projectdata. We navigated to our respective csvs, clicked on raw data to get the standalone csv, and saved it into a url variable in this Rmd. We then used the read_csv function from the readr library to import our data into rstudio. 
```{r import, message=FALSE}

mpx_cases_url = "https://raw.githubusercontent.com/PHW290/phw251_projectdata/main/euro_mpx_cases.csv"
mpx_cases_df <- read_csv(mpx_cases_url)

pop_denoms_url = "https://raw.githubusercontent.com/PHW290/phw251_projectdata/main/euro_pop_denominators.csv"
pop_denoms_df = read_csv(pop_denoms_url)

census_stats_url = "https://raw.githubusercontent.com/PHW290/phw251_projectdata/main/euro_census_stats.csv"
census_stats_df <- read_csv(census_stats_url,
          col_names = TRUE,
          col_types = NULL,
          na = c("", "NA"))

```