---
title: 'Final Report: Monkeypox infection rates in the European Union'
author: "Abigail Duque, Misty Garcia, Katherine Boudreau"
date: '2022-12-12'
output:
  html_document:
    df_print: paged
  pdf_document: default
subtitle: 'Milestone #6'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(tidyverse)
library(kableExtra)
library(plotly)
```

## Problem Statement 
We have been tasked with developing an update on Monkeypox (MPX) for the leadership of our state health department. MPX is a rare disease caused by infection with the monkeypox virus. As there has been a recent MPX outbreak in Europe, we will be analyzing their data to prepare a response for an outbreak here. 

We will be investigating how case rates changed over time in four different regions in Europe. We will also look at certain demographics and MPX rates by country. We will look at whether gender (male and female) affects case rates, as well as age (people 85 and older, and people younger than 85).


## Methods

We used four different datasets to complete this report:

  1.    Monkeypox data from the European Centre for Disease Prevention and Control (ECDC) 
  \href {https://www.ecdc.europa.eu/en/publications-data/data-monkeypox-cases-eueea}.     
  2.    European population denominators data from Eurostat
  \href{https://ec.europa.eu/eurostat/databrowser/view/tps00001/default/table?lang=en}.     
  3.    World regions and country codes from the International Organization for Standardization (ISO) 3166 (Geneva, Switzerland)
  \href{https://en.wikipedia.org/wiki/ISO_3166, https://www.iso.org/iso-3166-country-codes.html}.      
  4.    Demographic data from the European Statistical System's 2011 European census data
  \href{https://raw.githubusercontent.com/PHW290/phw251_projectdata/main/euro_census_stats.csv}.      

From the ISO 3166 world region data, we filtered by European region and selected the four European subregions and country codes. We then matched country code formats with the ECDC MPX & Eurostat population datasets in preparation for joining. The ECDC MPX dataset spanned 2022-05-09 to 2022-08-23 and included the reported date of confirmed monkeypox cases by country and the source of data collection. From this dataset, we filtered by date, country code and confirmed number of MPX cases. To obtain the population denominators used in our calculation of weekly MPX case rates, we incorporated Eurostat's European population dataset, which included yearly population data of European countries from 2011 to 2022 based on the total population residing in that country as of January 1st of each year. We used these three subsetted datasets for our investigations, MPX case rates by European subregions and MPX case rates by demographic factors.

To calculate MPX case rates by European subregion, we first filtered the European population dataset by the year 2022 (the only year ECDC MPX cases were recorded) and then selected country code and total population. We then inner joined the European population denominators, ECDC MPX and ISO3166 subsetted datasets together by country code. There were no discrepancies between datasets, therefore, no missing values after joining. Our final aggregated dataset contained five columns: European sub region, country code, confirmed number of monkeypox cases and population denominators data. After grouping by the four European subregions, we created a new MPX case rate by one million variable which can be visualized in Figure 1.

The European Statistical System's 2011 European census data contained EU country codes, sex at birth, age ranges, employment status, education status, population of locality of residence and the number of people in each strata. Since we were only interested in observing possible trends in European countries' MPX case rates when stratified by the female gender and people aged 85 years and older, we created new dataframes grouped by country code and filtered for the female sex and people aged 85 years and older per country followed by subsequent calculation of percentages of each respective demographic per country. We then joined our aggregated census data with our data frame containing MPX case rates by European country (Table 1). Trends in European MPX case rates by people 85 years of age and older and female gender can be seen in Figure 2 and Figure 3, respectively.

```{r Milestone 1, message=FALSE, echo=FALSE, warning=FALSE }
## Milestone 1

#Github link: https://github.com/r-public-health-group/monkey-pox
#Team roles and responsibilities: https://github.com/r-public-health-group/monkey-pox/blob/main/team_roles_and_responsibilities.docx
```

```{r Milestone 2  dataset import, message=FALSE, echo=FALSE, warning=FALSE}
## Milestone 2

### What is the data source?
#We are interested in how monkeypox case rates differ by European region and population demographics. We are first utilizing a monkeypox case dataset from the European Centre for Disease Prevention and Control that spans 2022-05-09 to 2022-08-23 and includes the daily number of confirmed monkey pox cases by country and how the data was collected. Then, we will incorporate a European population dataset from \href{https://ec.europa.eu/eurostat/databrowser/view/tps00001/default/table?lang=en}{Eurostat}, which includes European countries' yearly population data from 2011 to 2022 based on the total population residing in that country as of January 1st of each year. Finally, we will incorporate the European Statistical System's 2011 European census data which includes the EU country codes, sex at birth, age ranges, employment status (CAS), education status, population of locality of residence and the number of people in each strata.

### How does the dataset relate to the group problem statement and question?
#As we prepare our report on monkeypox case rates to the state health department, the monkeypox case data will provide us with the number of cases that will be the numerator of our case rate calculation, the population data gives us the size of the population at risk that will be the denominator of our case rate calculation, and the census data will allow us to stratify the case rates by various demographic factors.

### Import Statement
#All four of our CSVs are hosted on the course github page: https://github.com/PHW290/phw251_projectdata. We navigated to our respective csvs, clicked on raw data to get the standalone csv, and saved it into a url variable in this Rmd. We then used the read_csv function from the readr library to import our data into rstudio. 

mpx_cases_url = "https://raw.githubusercontent.com/PHW290/phw251_projectdata/main/euro_mpx_cases.csv"
mpx_cases_df <- read_csv(mpx_cases_url)

pop_denoms_url = "https://raw.githubusercontent.com/PHW290/phw251_projectdata/main/euro_pop_denominators.csv"
pop_denoms_df = read_csv(pop_denoms_url)

census_stats_url = "https://raw.githubusercontent.com/PHW290/phw251_projectdata/main/euro_census_stats.csv"
census_stats_df <- read_csv(census_stats_url,
          col_names = TRUE,
          col_types = NULL,
          na = c("", "NA"))

regions <- read.csv('https://raw.githubusercontent.com/PHW290/phw251_projectdata/main/world_country_regions.csv')
```

```{r lowercase, echo=FALSE, warning=FALSE}
### Lowercase Column Names
#We are now lowercasing all the column names.  
mpx_cases_df <- rename_with(mpx_cases_df, ~tolower(.x))
pop_denoms_df <- rename_with(pop_denoms_df, ~tolower(.x))
census_stats_df <- rename_with(census_stats_df, ~tolower(.x))
```

```{r, warning=FALSE, include=FALSE}
### Data Types and Descriptions
## Details of key data elements are outlined below. All elements listed are in an appropriate data format for the joins and analysis we intend to do.
#### Variable: daterep from mpx_cases_df
str(mpx_cases_df$daterep)
min(mpx_cases_df$daterep)
max(mpx_cases_df$daterep)
unique(dplyr::count(mpx_cases_df, daterep)$n)
#The dates are in date format, they span from 2022-05-09 to 2022-08-23, and each date shows up 29 times. 
```

```{r, warning=FALSE, include=FALSE}
#### Variable: daterep from mpx_cases_df
typeof(mpx_cases_df$countrycode)
unique(mpx_cases_df$countrycode)
length(unique(mpx_cases_df$countrycode))
#The variable countryexp is the country name where countrycode is it's two letter counterpart. Both are in character format. There are 29 countries respresented. 
```

```{r, warning=FALSE, include=FALSE}
#### Variable: source from mps_cases_df
typeof(mpx_cases_df$source)
unique(mpx_cases_df$source)
#Source is in a character format. There are two sources of data: "TESSy" and "EI."
```

```{r, warning=FALSE, include=FALSE}
#### Variable: confcase from mpx_cases_df
typeof(mpx_cases_df$confcases)
unique(mpx_cases_df$confcases)
min(mpx_cases_df$confcases)
mean(mpx_cases_df$confcases)
max(mpx_cases_df$confcases)
#The confirmed cases are in a numeric data format and range from 0 to 655. The mean is 5.7 cases per day.
```

```{r pop_denoms_df$geo, warning=FALSE, include=FALSE}
#### Variable: geo from pop_denoms_df
typeof(pop_denoms_df$geo)
length(unique(pop_denoms_df$geo))
#The values in the GEO column of the pop_denoms_df are characters denoting the geopolitical region that a given row's population data comes from. There are 54 possible values for this column, either a 2-letter country code or a code denoting the entire European Union on a given year. We will eventually merge these 2-letter country codess with the 2-letter country codes in the euro_mpx_cases csv, so the character data type will suffice.
```

```{r pop_denoms_df$obs_value, warning=FALSE, include=FALSE}
#### Variable: obs_value from pop_denoms_df
summary(pop_denoms_df$obs_value)
typeof(pop_denoms_df$obs_value)
#The values in the obs_value column are the observed population on January 1 of the reported year. These values are doubles representing the total population count of a given region, so we can use these values directly as the denominators in our case rate calculations.
```

```{r country_code, warning=FALSE, include=FALSE}
#### Variable: country_code from census_stats_df
typeof(census_stats_df$country_code)
unique(census_stats_df$country_code)
length(unique(census_stats_df$country_code))
#Country Code is a character data type and there are 32 different two-letter country codes in this dataset. We will obtain the different regions from the country_code variable.
```

```{r sex, warning=FALSE, include=FALSE}
#### Variable: sex from census_stats_df
unique(census_stats_df$sex)
#Sex is a character data type, and there are two categories: male, female. We might use sex as a demographic variable to stratify case rates by.
```

```{r age, warning=FALSE, include=FALSE}
#### Variable: age from census_stats_df
typeof(census_stats_df$age)
unique(census_stats_df$age)
#Age is character data type that includes six different age ranges in the dataset: <15, 15-29, 30-49, 65-84 & >85. We might use age as a demographic variable to stratify case rates by.
```

```{r employment status, warning=FALSE, include=FALSE}
#### Variable: cas from census_stats_df
typeof(census_stats_df$cas)
unique(census_stats_df$cas)
#Cas is a character data type and represents the employment status of an individual. There are five different statuses, and we might use these to stratify case rates by.
```

```{r education status, warning=FALSE, include=FALSE}
#### Variable: edu from census_stats_df
typeof(census_stats_df$edu)
unique(census_stats_df$edu)
#Education status is in a character format, and there are nine different statuses of education. We might use education status to stratify case rates by.
```

```{r residence population, warning=FALSE, include=FALSE}
#### Variable: res_pop from census_stats_df
typeof(census_stats_df$res_pop)
unique(census_stats_df$res_pop)
#Resident population (res_pop) is in a character data format. There are seven levels of population in each locality of residence.
```

```{r # of people in each location strata, warning=FALSE, include=FALSE}
#### Variable: pop from census_stats_df
typeof(census_stats_df$pop)
range(census_stats_df$pop)
#The strata population is in a numeric data format. The number of people in the different location strata ranges from 0 to 1,702,270.
```

```{r subset, warning=FALSE, include=FALSE}
## Milestone 3

#Subset rows or columns, as needed
#filtered by europe countries
regions <- regions %>%
  filter(region == 'Europe') %>%
  select(iso_3166.2, sub.region, region)

#create country code variable to join tables
regions <- regions %>% 
  mutate(country_code = substr(regions$iso_3166.2,12,13)) %>%
  select(-iso_3166.2,)

#filtered mpx_cases columns
mpx_cases_df <- mpx_cases_df %>%
  select(daterep, countrycode, confcases)

#filter pop_denom for 2011 for census analysis
pop_denoms_2011 <- pop_denoms_df %>%
  filter(time_period == 2011) %>%
  select(geo, obs_value)

#filtered pop_denom by year and columns
pop_denoms_df <- pop_denoms_df %>%
  filter(time_period == 2022) %>%
  select(geo, obs_value)

#joined all three datasets together
df <- inner_join(x = regions, y = mpx_cases_df, by = c('country_code' = 'countrycode'))
df <- inner_join(x = df, y = pop_denoms_df, by = c('country_code' = 'geo'))

df <- df %>%
  select(-region)
```

```{r new_variables, warning=FALSE, include=FALSE}
#Create new variables needed for analysis (minimum 2)

#created case rate by one million people
df_grouped <- df %>% 
  group_by(sub.region, daterep) %>%
  summarise(total_cases = sum(confcases), total_obs = sum(obs_value)) %>%
  mutate(case_rate_by_onemillion = total_cases / total_obs * 1000000)

```

``` {r clean_variables, warning=FALSE, include=FALSE}

#we used an inner join and there were no discrepancies between datasets, 
#therefore are no nulls caused by joining
sum(is.na(df_grouped))

#make sub.region a categorical value
df_grouped <- df_grouped %>%
  mutate(sub.region = factor(sub.region))

#make new month column & made it a categorical value
df_grouped <- df_grouped %>%
  mutate(month = strftime(daterep,"%B")) %>%
  mutate(month = factor(month, levels = c('May','June','July','August')))

str(df_grouped)

```

```{r warning=FALSE, include=FALSE}
#One or more tables with descriptive statistics for 4 data element
summary(df_grouped, echo=FALSE, warning=FALSE)
```

``` {r join census data, warning=FALSE, include=FALSE}

### Prepare census data

#To evaluate how census data trends for a region compare to that region's monkeypox rates, we will first calculate the % of the population in each region that is 85 years old or older and the % of the population in each region that is female. Then we will compare these rates against the regions' monkey pox rates and evaluate any trends we see.

# First, calculate the number of females in each country.
sex_pct_df <- census_stats_df %>%
  group_by(country_code, sex) %>%
  summarize(sex_count = sum(pop)) %>%
  group_by(country_code) %>%
  mutate(country_pop = sum(sex_count)) %>%
  filter(sex == "F",
         country_pop != 0) %>%
  rename(female_count = sex_count)

# Then, calculate the number greater than or equal to 85 years old in each country.
age_pct_df <- census_stats_df %>%
  group_by(country_code, age) %>%
  summarize(age_count = sum(pop)) %>%
  group_by(country_code) %>%
  mutate(country_pop = sum(age_count)) %>%
  filter(age == "Y_GE85", 
         country_pop != 0) %>%
  rename(ge85_count = age_count)

# Join these two df's together.
demog_df <- full_join(x = sex_pct_df, y = age_pct_df, by = "country_code") 
demog_df <- demog_df %>%
  select(country_code, ge85_count, female_count, country_pop.x) %>%
  rename(country_pop = country_pop.x)

# Join the combined df with the regions df so we can map country codes to regions.
##demog_df <- inner_join(x = regions, y = demog_df, by = 'country_code')

# Calculate the % of the population that is GE85 and the % that is Female
demog_df <- demog_df %>%
  group_by(country_code) %>%
  summarize(pct_ge85 = round(sum(ge85_count) / sum(country_pop) * 100,2),
            pct_female = round(sum(female_count) / sum(country_pop) * 100, 2))

# Group the monkeypox case rate df by region and calculate regional case rate
df_case_rate <- df %>%
  group_by(country_code) %>%
  summarize(total_cases = sum(confcases),
            total_obs = sum(obs_value)) %>%
  mutate(case_rate_p_mill = total_cases / total_obs * 1000000)

# Join case rates with the regional census data
demog_df <- inner_join(x = demog_df, y = df_case_rate, by = "country_code")
demog_df <- select(demog_df, c(-total_cases, -total_obs))

```
```{r data dictionary, include=FALSE}

#Data dictionary based on clean dataset (minimum 4 data elements), including:
#Fields
#Variable name
#Data type
#Description

#sub.region: factor, region in europe (north, south, east, west)
#daterep: date, year-month-day by week from may to august 2022
#total_cases: num, total monkeypox cases in a subregion by week
#total_obs: num, total population by subregion
#case_rate_by_onemillion: num, the rate of monkeypox in a EU subregion by one million people
#month: date, the month from may to august 2022
#pct_ge85: num, percent of the population aged 85 years or older
#pct_female: num, percent of the population that is female
#case_rate_p_mill: num, the rate of monkeypox in a European country by a million people
#female_count: num, number population that is female
#ge85_count: num, number of population that is aged 85 years or older
#country_pop: num, the number of people in the population for a European country

```

## Results    
  
  
### Figure 1

```{r Milestone 4, warning=FALSE, echo=FALSE}

plot_ly(
  df_grouped,
  x= ~month,
  y= ~case_rate_by_onemillion,
  color= ~sub.region,
  type="bar"
) %>%
  layout(
    title="MPX Case Rates by EU Region, May to August 2022",
    yaxis=list(title="Case Rates per 1,000,000"),
    xais=list(title="Month"),
    paper_bgcolor="white",
    plot_bgcolor="white"
  )

```
Here, one can observe a bar chart representing the case rates per 1,000,000 individuals across the four regions of Europe during May, June, July, and August of 2022.

All four regions experienced a similar chronological trend from May 2022 to August 2022 where May had the lowest case rates and July had the highest case rates. Between the four regions, there was a consistent ordering of case rates at each time point: Eastern Europe had the lowest rate, followed by Northern Europe, Western Europe, and Southern Europe. One interesting anomaly is that Northern Europe does not appear to experience as significant of a downward trend from July to August that the other regions experienced. 


### Table 1   


``` {r, warning=FALSE, echo=FALSE}
kable(demog_df,longtable=T,booktabs=T,
      col.names=c("Country Code", "% 85 and Older", "% Female", "Case Rate per Million" ),
      caption="2011 Census Rates versus 2022 Monkeypox Rates by Country") %>%
  kable_styling(full_width=F) %>%
  kable_styling(position="center") %>%
  kable_styling(font_size=14)
```
  
We compiled this table to investigate if a country's demographic data - in this table, age proportion of the population that is elderly or female - correlates with that country's monkeypox case rate.

From this table view, there appears to be a positive trend between percent of the population that is 85 and older and that country's monkeypox case rate. A trend is not obvious between a country's percent of the population that is female and its monkeypox rate, especially because the proportion of the population that is female is very similar across all the countries. Plotting these data longitudinally, as we do below, will help to confirm these observations.


### Figure 2    

  
```{r case rate vs 85yo plot, warning=FALSE, echo=FALSE}
#Creating a visual plot will help elucidate the demographic relationship better.

ggplot(demog_df, aes(pct_ge85, case_rate_p_mill)) +
  geom_point() + 
  geom_smooth(formula = y ~ x, method = lm, se=FALSE) + 
  labs(x='Percent of the country that is 85 years and older',
       y='Case rate of country per 1,000,000',
       title='Monkeypox rates increase in countries with more people who are 85 and older') +
  theme_light()
```

In Figure 2, we plot the percent of a country's population that is aged 85 years and older against the MPX case rates per 1,000,000 individuals in European countries, and we observe a fairly strong positive trend, confirming a positive correlation between these two variables for these countries.
  
### Figure 3  


```{r case rate vs females plot, warning=FALSE, echo=FALSE}
ggplot(demog_df, aes(pct_female, case_rate_p_mill)) +
  geom_point() + 
  geom_smooth(formula = y ~ x, method=lm, se=FALSE) + 
    labs(x='Percent of the country that is female',
       y='Case rate of country per 1,000,000',
       title='Monkeypox rates decrease in countries with more females') +
  theme_light()
```
  
In Figure 3, we plot the percent of a country's population that is female against the MPX case rates per 1,000,000 individuals in European countries, and we see a negative correlation between these two variables.
  
## Discussion 
In conclusion, monkeypox case rates followed a similar trend across all European regions from May to August 2022, with the lowest case rates occurring in May of 2022, and the highest case rates occurring in July of 2022. Homing in on the European countries' demographic data we further see trends between country case rate data and population demographic data for two key demographics: proportion of the population that is 85 and older and proportion of the population that is female. For the former we see a positive correlation and for the latter we see a negative correlation.

We hope these trends can help public health leaders understand and prioritize monkeypox interventions.
